A Sound and Complete Deep Embedding of Arbitrary Normal Modal Logics in Coq

### 2024-01-10 paper/COLA/paper.tex
Adding paper to COLA

%%
%% This is file `sample-sigconf.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `sigconf')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigconf.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigconf]{acmart}

%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\copyrightyear{2022}
\acmYear{2022}
\setcopyright{acmcopyright}\acmConference[SBLP 2022]{XXVI Brazilian Symposium
on Programming Languages}{October 6--7, 2022}{Virtual Event, Brazil}
\acmBooktitle{XXVI Brazilian Symposium on Programming Languages (SBLP 2022),
October 6--7, 2022, Virtual Event, Brazil}
\acmPrice{15.00}
\acmDOI{10.1145/3561320.3561329}
\acmISBN{978-1-4503-9744-5/22/10}

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

%%
%% For managing citations, it is recommended to use bibliography
%% files in BibTeX format.
%%
%% You can then either use BibTeX with the ACM-Reference-Format style,
%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
%% support for advanced citation of software artefact from the
%% biblatex-software package, also separately available on CTAN.
%%
%% Look at the sample-*-biblatex.tex files for templates showcasing
%% the biblatex styles.
%%

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
%%\citestyle{acmauthoryear}

\usepackage{minted}
\usepackage{bussproofs}
\usepackage{multirow}

%%
%% end of the preamble, start of the body of the document source.
\begin{document}
%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{A Sound Deep Embedding of Arbitrary Normal Modal Logics in Coq}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.

\author{Ariel Agne da Silveira}
\affiliation{\institution{Federal University of Ouro Preto}\country{}}
\email{ariel.silveira@aluno.ufop.edu.br}

\author{Miguel Nunes}
\affiliation{\institution{Santa Catarina State University}\country{}}
\email{miguel.alfredo.nunes@gmail.com}

\author{Karina Roggia}
\affiliation{\institution{Santa Catarina State University}\country{}}
\email{karina.roggia@udesc.br}

\author{Paulo Torrens}
\affiliation{\institution{Santa Catarina State University}\country{}}
\email{paulotorrens@gnu.org}

\author{Rodrigo Ribeiro}
\affiliation{\institution{Federal University of Ouro Preto}\country{}}
\email{rodrigo.ribeiro@ufop.edu.br}

%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
\renewcommand{\shortauthors}{da Silveira et al.}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  This work describes an encoding of modal logics using the Coq proof assistant.
  Our formalization differs from previous attempts by providing a uniform
  representation of several systems for modal logic using Coq's rich type
  structure. We illustrate the usefulness of our library in a formalization of
  L\"ob's theorem which closely follows a classical proof of this result.
\end{abstract}

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{Modal logics, soundness, Coq proof assistant}

\maketitle

@@

\begin{itemize}
  \item We define a \emph{deep embedding} of modal logic syntax in Coq and
    define its Kripke semantics.
  \item We model the Hilbert deductive system as a Coq inductive type that is
    parameterized by different axiom systems like K, D, B, T, K4, K5, S4, S5 and GL,
    and prove its soundness with respect to Kripke semantics.
    In such, we also formalize several frame correspondence results.
  \item As an example of usage, we show a proof of L\"ob's theorem using the deductive system
    developed in our library.
\end{itemize}

The rest of this work is organized as follows: Section~\ref{sec:background} provides
brief introduction on modal logics and the Coq proof assistant. Details about
the implementation of our Coq library are presented in
Section~\ref{sec:implementation}. A case study in proving L\"ob's theorem is
the subject of Section~\ref{sec:case}. Related works are discussed
on Section~\ref{sec:related} and Section~\ref{sec:conclusion} draws some conclusions and
provide some pointers for future work.

@@

\paragraph{An overview of Coq proof assistant.}

@@

However, writing a proof term whose type is that of a logical formula
can be challenging, even for very simple propositions.  In order to
make the writing of complex proofs easier, Coq provides
\emph{tactics}, which are commands that can be used to construct proof
terms in a more user-friendly way.

As a tiny example, consider the task of proving the following 
formula of the propositional logic:
\[
(A \to B)\to (B\to C) \to A \to C
\]
In Coq, such theorem can be expressed as:
\begin{minted}{coq}
Section EXAMPLE.
   Variables A B C: Prop.
   Theorem example: (A -> B) -> (B -> C) -> A -> C.
   Proof.
       intros H H' HA.
       apply H'.
       apply H.
       assumption. 
   Qed.
End EXAMPLE.
\end{minted}
In the previous source code piece, we have defined a Coq section named
\mintinline{coq}{EXAMPLE}\footnote{In Coq, we can use sections to delimit the
  scope of local variables.} which declares variables \mintinline{coq}{A},
\mintinline{coq}{B} and \mintinline{coq}{C} as being propositions (i.e. with type
\mintinline{coq}{Prop}). Tactic \mintinline{coq}{intros} introduces variables
\mintinline{coq}{H}, \mintinline{coq}{H'} and \mintinline{coq}{HA} into the (typing) context,
respectively with types \mintinline{coq}{A -> B}, \mintinline{coq}{B -> C} and
\mintinline{coq}{A} and leaves goal \mintinline{coq}{C} to be proved. Tactic
\mintinline{coq}{apply}, used with a term \mintinline{coq}{t}, generates goal
\mintinline{coq}{P}
when there exists \mintinline{coq}{t: P -> Q} in the typing context and the
current goal is \mintinline{coq}{Q}. Thus, \mintinline{coq}{apply H'} changes the goal
from \mintinline{coq}{C} to \mintinline{coq}{B} and \mintinline{coq}{apply H}
changes the goal to \mintinline{coq}{A}. Tactic \mintinline{coq}{assumption}
traverses the typing context to find a hypothesis that matches the goal.

We define next a proof of the previous propositional logical formula
that, in contrast to the previous proof, that was built using tactics
(\mintinline{coq}{intros}, \mintinline{coq}{apply} and \mintinline{coq}{assumption}), is coded
directly as a function:
\begin{minted}{coq}
Definition example: (A -> B) -> (B -> C) -> A -> C :=
  fun (H: A -> B) (H': B -> C) (HA: A) => H' (H HA).
\end{minted}
However, even for very simple theorems, coding a definition directly
as a Coq term can be a hard task. Because of this, the use of tactics
has become the standard way of proving theorems in Coq. Furthermore,
the Coq proof assistant provides not only a great number of tactics
but also a domain-specific language for scripted proof automation,
called $\mathcal{L}$tac. Details about $\mathcal{L}$tac and Coq can be found
in~\cite{Chlipala13,Bertot10,manual_coq}.

\paragraph{An overview of modal logic.} Modal logics are designed for reasoning
about truth across various --- abstract --- worlds. In such logics, a
proposition may be true in some world, but false in another one. The systems of
modal logic considered in this work extend traditional propositional logic
with two operators on propositions: $\Box$ (box) and $\Diamond$ (diamond).
The syntax for modal logic formulas is defined by the following context free
grammar (where
$\mathcal{P} = \{A,B,C,...\}$ is the countable set of propositional variables):
\[
  \begin{array}{lcl}
    \varphi & ::= & \mathcal{P}\,|\,\neg\varphi\,|\,\varphi\land\varphi\,|\,\varphi\lor\varphi\,|\,\varphi\to\varphi\,|\,\Box\varphi\,|\,\Diamond\varphi
  \end{array}  
\]
Intuitively, the semantics of the operator $\Box$ is similar to an universal
quantifier, since it expresses \emph{necessity}: $\Box\,\varphi$ is true in the
current world iff $\varphi$ is true in every world accessible from the current
one. Similarly, the semantics of operator
$\Diamond$ is that a formula $\Diamond\varphi$ is true in the current
world iff $\varphi$ is \emph{possible}, i.e. true in at least some world
accessible from the current world. We the remaining of the paper, we also write
$\phi_1\leftrightarrow\phi_2$ to mean $\phi_1\to\phi_2\land\phi_2\to\phi_1$.

@@

\begin{figure}[H]
\begin{tabular}{cc}
    \AxiomC{$\Gamma\vdash\varphi_1\to\varphi_2$}
    \AxiomC{$\Gamma\vdash\varphi_1$}
    \RightLabel{MP}
    \BinaryInfC{$\Gamma\vdash\varphi_2$}
    \DisplayProof
  &
    \AxiomC{$\Gamma\vdash\varphi$}
    \RightLabel{Nec}
    \UnaryInfC{$\Gamma\vdash\Box \varphi$}
    \DisplayProof
\end{tabular}
\end{figure}
The set of axioms and the two previous rules form a sound and complete Hilbert-style
deductive system for modal logics~\cite{blackburn}.

@@

\begin{figure}[H]
\begin{tabular}{|c|c|}
  \hline
  Property of $R$                          & Formula schema \\ \hline
  reflexive: $\forall w.wRw$               & $\Box A \to A$ \\ 
  serial: $\forall w\exists w'. wRw'$      & $\Box A \to \Diamond A$ \\
  symmetric: $\forall w\,w'.wRw' \to w'Rw$ & $A \to \Box\Diamond A$ \\
  \hline
\end{tabular}
\end{figure}
There are other properties of interest for a relation $R$ like being
transitive, functional, euclidean, and so on. More about the correspondence
of frames and deductive systems properties can be found elsewhere~\cite{Chellas,blackburn}.

@@

\paragraph{Syntax and semantics of modal logics}
To implement our library, we use a \emph{deep embedding} of
formulas, i.e. we represent formulas as the following Coq inductive datatype.
\begin{minted}{coq}
Inductive formula: Set :=
| Lit    : nat -> formula
| Neg    : formula -> formula
| Box    : formula -> formula
| Dia    : formula -> formula
| And    : formula -> formula -> formula
| Or     : formula -> formula -> formula
| Implies: formula -> formula -> formula.
\end{minted}
Most constructors of the type have an immediate meaning, since they represent
connectives of propositional logic. The constructor \mintinline{coq}{Lit}
is used to construct propositional variables; constructor \mintinline{coq}{Box}
denotes the modal operator $\Box$ and \mintinline{coq}{Dia}, the operator
$\Diamond$. Furthermore, the type \mintinline{coq}{theory} is defined as a list
of formulas.

\begin{minted}{coq}
Definition theory := list formula.
\end{minted}

@@

\paragraph{The Hilbert deductive system}
Basically, a Hilbert-style deductive system allows the use of four
possibilities in a proof: 1) the use of a hypothesis; 2) the use of an axiom;
3) the use of the modus ponens rule or 4) the use of the generalization (necessity)
rule. The inductive type \mintinline{coq}{deduction} encodes these possibilities
as its constructors. The \mintinline{coq}{deduction} type is parameterized by
the set of axioms that form a system for some modal logic. In this way, our definition can be
used by several systems alike.
\begin{minted}{coq}
Inductive deduction (A: axiom -> Prop):
                          theory -> formula -> Prop :=
  (* Premise. *)
  | Prem: forall (t: theory)
                 (f: formula)
                 (i: nat),
          (nth_error t i = Some f) -> deduction A t f
  (* Axiom. *)
  | Ax: forall (t: theory)
               (a: axiom)
               (f: formula),
        A a -> instantiate a = f -> deduction A t f
  (* Modus Ponens. *)
  | Mp: forall (t: theory)
               (f g: formula)
               (d1: deduction A t (Implies f g))
               (d2: deduction A t f),
        deduction A t g
  (* Generalization. *)
  | Nec: forall (t: theory)
                (f: formula)
                (d1: deduction A t f),
         deduction A t (Box f).
\end{minted}
The constructors \mintinline{coq}{Prem}, \mintinline{coq}{Mp} and \mintinline{coq}{Nec}
represent rules for hypothesis, modus ponens and generalization, respectively.
Before discussing the \mintinline{coq}{Ax} constructor, we need to discuss how axioms are
represented in our library. The inductive type \mintinline{coq}{axiom} specifies
the different axioms supported by our library.
\begin{minted}{coq}
Inductive axiom: Set :=
  | ax1  : formula -> formula -> axiom
  | ax2  : formula -> formula -> formula  -> axiom
  (* some code omitted *)
  | axK  : formula -> formula -> axiom
  | axPos: formula -> formula -> axiom
  | axT  : formula -> axiom
  | axB  : formula -> axiom
  | axK4 : formula -> axiom
  | axD  : formula -> axiom
  | axK5 : formula -> axiom.
\end{minted}
Each constructor of type \mintinline{coq}{axiom} represents an axiom scheme and
its type stores the subformulas which compose the axiom. The type has
17 constructors: the first 10 encode axioms for propositional logic
and, the last 7 are for different formulas which are taken as axioms by common
systems of modal logic. Axioms are built using the function \mintinline{coq}{instantiate},
which receives an axiom and builds a corresponding formula by pattern matching as
follows:
\begin{minted}{coq}
Definition instantiate (a: axiom): formula :=
  match a with
  | ax1    p   q
     => [! p -> (q -> p) !]
  | ax2    p   q   r
     => [! (p -> (q -> r)) -> ((p -> q) -> (p -> r)) !]
  | ax3    p   q
     => [! (~ q -> ~ p) -> (p -> q) !]
  | axK    p   q
     => [! [] (p -> q) -> ([] p -> [] q) !]
  | axPos  p   q
     => [! <> (p \/ q) -> (<> p \/ <> q) !]
  (* some code omitted *)
  end.
\end{minted}
Notation \mintinline{coq}{[! p !]} is used to write modal logic formulas using a
syntax closer to mathematical practice using Coq. This is necessary to avoid
parsing ambiguities, since some symbols are also used by Coq's syntax.
Modalities $\Box$ and $\Diamond$ are denoted by \mintinline{coq}{[]}
and \mintinline{coq}{<>}, respectively.

@@

>>>>

% Template for Elsevier CRC journal article
% version 1.2 dated 09 May 2011

% This file (c) 2009-2011 Elsevier Ltd.  Modifications may be freely made,
% provided the edited file is saved under a different name

% This file contains modifications for Procedia Computer Science
% but may easily be adapted to other journals

% Changes since version 1.1
% - added "procedia" option compliant with ecrc.sty version 1.2a
%   (makes the layout approximately the same as the Word CRC template)
% - added example for generating copyright line in abstract

%-----------------------------------------------------------------------------------

%% This template uses the elsarticle.cls document class and the extension package ecrc.sty
%% For full documentation on usage of elsarticle.cls, consult the documentation "elsdoc.pdf"
%% Further resources available at http://www.elsevier.com/latex

%-----------------------------------------------------------------------------------

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                          %%
%% Important note on usage                                  %%
%% -----------------------                                  %%
%% This file should normally be compiled with PDFLaTeX      %%
%% Using standard LaTeX should work but may produce clashes %%
%%                                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% The '3p' and 'times' class options of elsarticle are used for Elsevier CRC
%% Add the 'procedia' option to approximate to the Word template
%\documentclass[3p,times,procedia]{elsarticle}
\documentclass[3p,times]{elsarticle}

%% The `ecrc' package must be called to make the CRC functionality available
\usepackage{ecrc}

%% The ecrc package defines commands needed for running heads and logos.
%% For running heads, you can set the journal name, the volume, the starting page and the authors

%% set the volume if you know. Otherwise `00'
\volume{00}

%% set the starting page if not 1
\firstpage{1}

%% Give the name of the journal:
\journalname{Journal of Computer Languages}

%% Give the author list to appear in the running head
%% Example \runauth{C.V. Radhakrishnan et al.}
\runauth{A.A. da Silveira et. al.}

%% The choice of journal logo is determined by the \jid and \jnltitlelogo commands.
%% A user-supplied logo with the name <\jid>logo.pdf will be inserted if present.
%% e.g. if \jid{yspmi} the system will look for a file yspmilogo.pdf
%% Otherwise the content of \jnltitlelogo will be set between horizontal lines as a default logo

%% Give the abbreviation of the Journal.  Contact the journal editorial office if in any doubt
\jid{COLA}

%% Give a short journal name for the dummy logo (if needed)
\jnltitlelogo{COLA}

%% Provide the copyright line to appear in the abstract
%% Usage:
%   \CopyrightLine[<text-before-year>]{<year>}{<restt-of-the-copyright-text>}
%   \CopyrightLine[Crown copyright]{2011}{Published by Elsevier Ltd.}
%   \CopyrightLine{2011}{Elsevier Ltd. All rights reserved}
\CopyrightLine{2011}{Published by Elsevier Ltd.}

%% Hereafter the template follows `elsarticle'.
%% For more details see the existing template files elsarticle-template-harv.tex and elsarticle-template-num.tex.

%% Elsevier CRC generally uses a numbered reference style
%% For this, the conventions of elsarticle-template-num.tex should be followed (included below)
%% If using BibTeX, use the style file elsarticle-num.bst

%% End of ecrc-specific commands
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
\usepackage{minted}
\usepackage{bussproofs}
\usepackage{multirow}
\usepackage{hyperref}
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers after \end{frontmatter}.
%% \usepackage{lineno}

%% natbib.sty is loaded by default. However, natbib options can be
%% provided with \biboptions{...} command. Following options are
%% valid:

%%   round  -  round parentheses are used (default)
%%   square -  square brackets are used   [option]
%%   curly  -  curly braces are used      {option}
%%   angle  -  angle brackets are used    <option>
%%   semicolon  -  multiple citations separated by semi-colon
%%   colon  - same as semicolon, an earlier confusion
%%   comma  -  separated by comma
%%   numbers-  selects numerical citations
%%   super  -  numerical citations as superscripts
%%   sort   -  sorts multiple citations according to order in ref. list
%%   sort&compress   -  like sort, but also compresses numerical citations
%%   compress - compresses without sorting
%%
%% \biboptions{comma,round}

% \biboptions{}

% if you have landscape tables
\usepackage[figuresright]{rotating}

% put your own definitions here:
%   \newcommand{\cZ}{\cal{Z}}
%   \newtheorem{def}{Definition}[section]
%   ...

% add words to TeX's hyphenation exception list
%\hyphenation{author another created financial paper re-commend-ed Post-Script}

% declarations for front matter

\begin{document}

\begin{frontmatter}

%% Title, authors and addresses

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for the associated footnote;
%% use the fnref command within \author or \address for footnotes;
%% use the fntext command for the associated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for the associated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%%
%% \title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
%% \author{Name\corref{cor1}\fnref{label2}}
%% \ead{email address}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
%% \address{Address\fnref{label3}}
%% \fntext[label3]{}

\dochead{}
%% Use \dochead if there is an article header, e.g. \dochead{Short communication}
%% \dochead can also be used to include a conference title, if directed by the editors
%% e.g. \dochead{17th International Conference on Dynamical Processes in Excited States of Solids}

\title{A Sound and Complete Deep Embedding of Arbitrary Normal Modal Logics in Coq}

%% use optional labels to link authors explicitly to addresses:
%% \author[label1,label2]{<author name>}
%% \address[label1]{<address>}
%% \address[label2]{<address>}

\author{Ariel Agne da Silveira}
\ead{ariel.silveira@aluno.ufop.edu.br}
\address{Federal University of Ouro Preto}

\author{Miguel Nunes}
\ead{miguel.alfredo.nunes@gmail.com}
\address{Santa Catarina State University}

\author{Karina Roggia}
\ead{karina.roggia@udesc.br}
\address{Santa Catarina State University}

\author{Paulo Torrens}
\ead{paulotorrens@gnu.org}
\address{Santa Catarina State University}

\author{Rodrigo Ribeiro}
\ead{rodrigo.ribeiro@ufop.edu.br}
\address{Federal University of Ouro Preto}

%
\begin{abstract}
  This work describes an encoding of modal logics using the Coq proof assistant.
  Our formalization differs from previous attempts by providing a uniform
  representation of several systems for modal logic using Coq's rich type
  structure. We illustrate the usefulness of our library in a formalization of
  L\"ob's theorem which closely follows a classical proof of this result.
\end{abstract}

\begin{keyword}
Modal logic \sep soundness \sep completeness \sep Coq proof assistant
\end{keyword}

\end{frontmatter}

%%
%% Start line numbering here if you want
%%
% \linenumbers

@@

\begin{itemize}
  \item We define a \emph{deep embedding} of modal logic syntax in Coq and
    define its Kripke semantics.
  \item We model the Hilbert deductive system as a Coq inductive type that is
    parameterized by different axiom systems like K, D, B, T, K4, K5, S4, S5 and GL.
  \item We formally verify several frame correspondence results.
  \item We prove soundness and completeness for the Hilbert deductive system. For the
        completeness proof we follow the so-called Henkin-style and it closely follows
        the proofs presented in classical modal logics textbooks~\cite{Blackburn-ModalLogics}.
  \item As an example of usage, we show a proof of L\"ob's theorem using the deductive system
    developed in our library.
\end{itemize}

This article is an extended version of our SBLP 2022 work~\cite{Silveira22}. In this paper we include
a detailed discussion about our formalization of a completeness proof for the Hilbert deductive
system following the strategies found in traditional modal logic books~\cite{Blackburn-ModalLogics}.
The rest of this work is organized as follows: Section~\ref{sec:background} provides
brief introduction on modal logics and the Coq proof assistant. Details about
the implementation of our Coq library are presented in
Section~\ref{sec:implementation}. A case study in proving L\"ob's theorem is
the subject of Section~\ref{sec:case}. Related works are discussed
on Section~\ref{sec:related} and Section~\ref{sec:conclusion} draws some conclusions and
provide some pointers for future work.

@@

\subsection{An overview of Coq proof assistant.}

@@

In order to make writing proofs easier, Coq provides
\emph{tactics}, which are commands that can be used to construct proof
terms in a more user-friendly way. As an example, consider the task of proving the following 
formula of the propositional logic:
\[
(A \to B)\to (B\to C) \to A \to C
\]
In Coq, such theorem can be expressed as:
\begin{minted}{coq}
Section EXAMPLE.
  Variables A B C: Prop.
  Theorem example: (A -> B) -> (B -> C) -> A -> C.
  Proof.
    intros H H' HA.
    apply H'.
    apply H.
    assumption. 
  Qed.
End EXAMPLE.
\end{minted}
In the above, we have defined a Coq section named
\mintinline{coq}{EXAMPLE}\footnote{In Coq, we can use sections to delimit the
  scope of local variables.} which declares variables \mintinline{coq}{A},
\mintinline{coq}{B} and \mintinline{coq}{C} as being propositions (i.e. with type
\mintinline{coq}{Prop}). Tactic \mintinline{coq}{intros} introduces variables
\mintinline{coq}{H}, \mintinline{coq}{H'} and \mintinline{coq}{HA} into the typing context,
respectively with types \mintinline{coq}{A -> B}, \mintinline{coq}{B -> C}, and
\mintinline{coq}{A}, and leaves goal \mintinline{coq}{C} to be proved. Tactic
\mintinline{coq}{apply}, used with a term \mintinline{coq}{t}, generates goal
\mintinline{coq}{P}
when there exists \mintinline{coq}{t: P -> Q} in the typing context and the
current goal is \mintinline{coq}{Q}. Thus, \mintinline{coq}{apply H'} changes the goal
from \mintinline{coq}{C} to \mintinline{coq}{B} and \mintinline{coq}{apply H}
changes the goal to \mintinline{coq}{A}. Tactic \mintinline{coq}{assumption}
traverses the typing context to find a hypothesis that matches the goal,
concluding the proof.

We define next a proof of the previous propositional logical formula
that, in contrast to the previous proof, that was built using tactics, is coded
directly as a function:
\begin{minted}{coq}
Definition example: (A -> B) -> (B -> C) -> A -> C :=
  fun (H: A -> B) (H': B -> C) (HA: A) => H' (H HA).
\end{minted}
However, even for very simple theorems, coding a definition directly
as a Coq term can be a hard task. Because of this, the use of tactics
has become the standard way of proving theorems in Coq. Furthermore,
the Coq proof assistant provides not only a great number of tactics
but also a domain-specific language for scripted proof automation,
called $\mathcal{L}$tac. Details about $\mathcal{L}$tac and Coq can be found
in~\cite{Chlipala13,Bertot10,manual_coq}.

\subsection{An overview of modal logic.} Modal logics are designed for reasoning
about truth across various --- abstract --- worlds. In such logics, a
proposition may be true in some world, but false in another one. The systems of
modal logic considered in this work extend traditional propositional logic
with two operators on propositions: $\Box$ (box) and $\Diamond$ (diamond).
The syntax for modal logic formulas is defined by the following context free
grammar (where
$\mathcal{P} = \{A,B,C,...\}$ is the countable set of propositional variables):
\[
  \begin{array}{lcl}
    \varphi & ::= & \mathcal{P}\,|\,\neg\varphi\,|\,\varphi\land\varphi\,|\,\varphi\lor\varphi\,|\,\varphi\to\varphi\,|\,\Box\varphi\,|\,\Diamond\varphi
  \end{array}  
\]
Intuitively, the semantics of the operator $\Box$ is similar to an universal
quantifier, since it expresses \emph{necessity}: $\Box\,\varphi$ is true in the
current world iff $\varphi$ is true in every world accessible from the current
one. Similarly, the semantics of operator
$\Diamond$ is that a formula $\Diamond\varphi$ is true in the current
world iff $\varphi$ is \emph{possible}, i.e. true in at least some world
accessible from the current world. We the remaining of the paper, we also write
$\phi_1\leftrightarrow\phi_2$ to mean $\phi_1\to\phi_2\land\phi_2\to\phi_1$.

@@

\begin{figure}[H]
\begin{tabular}{cc}
    \AxiomC{$\Gamma\vdash\varphi_1\to\varphi_2$}
    \AxiomC{$\Gamma\vdash\varphi_1$}
    \RightLabel{MP}
    \BinaryInfC{$\Gamma\vdash\varphi_2$}
    \DisplayProof
  &
    \AxiomC{$\Gamma\vdash\varphi$}
    \RightLabel{Nec}
    \UnaryInfC{$\Gamma\vdash\Box \varphi$}
    \DisplayProof
\end{tabular}
\centering
\end{figure}
The set of axioms and the two previous rules form a sound and complete Hilbert-style
deductive system for modal logics~\cite{blackburn}. Given a formula $\varphi$ and a
set of hypothesis $\Gamma$, the soundness property ensures that
whenever $\varphi$ is deduced from $\Gamma$, then $\Gamma\models \varphi$. On the
other hand, completeness guarantee that if $\Gamma\models\varphi$, then $\varphi$ is
provable from $\Gamma$ using the deductive system. While the soundness argument can
be proved by a immediate inductive argument over derivations of $\Gamma\vdash\varphi$,
the same strategy does not apply to the completeness property: the key step in proving
completeness is to produce a \emph{cannonical model} using maximal consistent sets
of formulas.

@@

\begin{figure}[H]
\begin{tabular}{|c|c|}
  \hline
  Property of $R$                          & Formula schema \\ \hline
  reflexive: $\forall w.wRw$               & $\Box A \to A$ \\ 
  serial: $\forall w\exists w'. wRw'$      & $\Box A \to \Diamond A$ \\
  symmetric: $\forall w\,w'.wRw' \to w'Rw$ & $A \to \Box\Diamond A$ \\
  \hline
\end{tabular}
\centering
\end{figure}
There are other properties of interest for a relation $R$ like being
transitive, functional, euclidean, and so on. More about the correspondence
of frames and deductive systems properties can be found elsewhere~\cite{Chellas,blackburn}.

@@

\subsection{Syntax and semantics of modal logics}

To implement our library, we use a \emph{deep embedding} of
formulas, i.e. we represent formulas as the following Coq inductive datatype.
\begin{minted}{coq}
Inductive formula: Set :=
| Lit    : nat -> formula
| Neg    : formula -> formula
| Box    : formula -> formula
| Dia    : formula -> formula
| And    : formula -> formula -> formula
| Or     : formula -> formula -> formula
| Implies: formula -> formula -> formula.
\end{minted}
Most constructors of the type have an immediate meaning, since they represent
connectives of propositional logic. The constructor \mintinline{coq}{Lit}
is used to construct propositional variables; constructor \mintinline{coq}{Box}
denotes the modal operator $\Box$ and \mintinline{coq}{Dia}, the operator
$\Diamond$. Furthermore, the type \mintinline{coq}{theory} is defined as a function
between \mintinline{coq}{formula} and Coq's \mintinline{coq}{Prop}s.

\begin{minted}{coq}
Definition theory := formula -> Prop.
\end{minted}

@@

\subsection{The Hilbert deductive system}

Basically, a Hilbert-style deductive system allows the use of four
possibilities in a proof: 1) the use of a hypothesis; 2) the use of an axiom;
3) the use of the modus ponens rule or 4) the use of the generalization (necessity)
rule. The inductive type \mintinline{coq}{deduction} encodes these possibilities
as its constructors. The \mintinline{coq}{deduction} type is parameterized by
the set of axioms that form a system for some modal logic. In this way, our definition can be
used by several systems alike.
\begin{minted}{coq}
Inductive deduction (A: axiom -> Prop):
                          theory -> formula -> Prop :=
  (* Premise. *)
  | Prem: forall (t: theory)
                 (f: formula),
        t f -> deduction A t f
  (* Axiom. *)
  | Ax: forall (t: theory)
               (a: axiom)
               (f: formula),
        A a -> instantiate a = f -> deduction A t f
  (* Modus Ponens. *)
  | Mp: forall (t: theory)
               (f g: formula)
               (d1: deduction A t (Implies f g))
               (d2: deduction A t f),
        deduction A t g
  (* Generalization. *)
  | Nec: forall (t: theory)
                (f: formula)
                (d1: deduction A t f),
         deduction A t ([] f).
\end{minted}
The constructors \mintinline{coq}{Prem}, \mintinline{coq}{Mp} and \mintinline{coq}{Nec}
represent rules for hypothesis, modus ponens and generalization, respectively.
Before discussing the \mintinline{coq}{Ax} constructor, we need to discuss how axioms are
represented in our library. The inductive type \mintinline{coq}{axiom} specifies
the different axioms supported by our library.
\begin{minted}{coq}
Inductive axiom: Set :=
  | ax1  : formula -> formula -> axiom
  | ax2  : formula -> formula -> formula  -> axiom
  (* some code omitted *)
  | axK  : formula -> formula -> axiom
  | axPos: formula -> formula -> axiom
  | axT  : formula -> axiom
  | axB  : formula -> axiom
  | axK4 : formula -> axiom
  | axD  : formula -> axiom
  | axK5 : formula -> axiom.
\end{minted}
Each constructor of type \mintinline{coq}{axiom} represents an axiom scheme and
its type stores the subformulas which compose the axiom. The type has
17 constructors: the first 10 encode axioms for propositional logic
and, the last 7 are for different formulas which are taken as axioms by common
systems of modal logic. Axioms are built using the function \mintinline{coq}{instantiate},
which receives an axiom and builds a corresponding formula by pattern matching as
follows:
\begin{minted}{coq}
Definition instantiate (a: axiom): formula :=
  match a with
  | ax1    p   q
     => [! p -> (q -> p) !]
  | ax2    p   q   r
     => [! (p -> (q -> r)) -> ((p -> q) -> (p -> r)) !]
  | ax3    p   q
     => [! (~ q -> ~ p) -> (p -> q) !]
  | axK    p   q
     => [! [] (p -> q) -> ([] p -> [] q) !]
  | axPos  p   q
     => [! <> (p \/ q) -> (<> p \/ <> q) !]
  (* some code omitted *)
  end.
\end{minted}
Notation \mintinline{coq}{[! p !]} is used to write modal logic formulas using a
syntax closer to mathematical practice using Coq. This is necessary to avoid
parsing ambiguities, since some symbols are also used by Coq's syntax.
Modalities $\Box$ and $\Diamond$ are denoted by \mintinline{coq}{[]}
and \mintinline{coq}{<>}, respectively.

@@

%%
%% End of file `ecrc-template.tex'.
